---
title: "Exploratory data analysis and PCA"
#author:
output:
  #rmdformats::html_clean:
  rmdformats::downcute:
    downcute_theme: "chaos"
    use_bookdown: false
    self_contained: true
    thumbnails: true
    lightbox: true
    highlight: tango
    gallery: true
    toc_depth: 3
    code_folding: show
    css: css/custom.css
knit:
  (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding =
  encoding, output_dir = "html") })
---

**!! The script works in R, however cannot be knitted in markdown because of an unresolved issue !!**
```
Error: Failed to open Data/ADAPTmap_genotypeTOP_20160222_full/out/plinkMDS_filtered_goats.log.  Try changing the --out parameter.
```

```{r}
rm(list = ls())
getwd()
list.files()
```
```{r}
library(tidyverse)
library(patchwork)
library(ggsci)
library(ggdark)
knitr::opts_chunk$set(error = TRUE)
```
We start from the data after QC of the file `2_QC_plink.Rmd`.

We compute the IBS (Identity-By-State) distance between individuals.
Not Euclidean distances!

`--distance 1-ibs flat-missing` replaces `--distance-matrix` in PLINK 1.9.

```{r}
# system("plink --distance --help")
system("plink \\
--chr-set 29 no-xy \\
--allow-no-sex \\
--nonfounders \\
--file Data/ADAPTmap_genotypeTOP_20160222_full/out/ADAPTmap_TOP \\
--distance-matrix \\
--out Data/ADAPTmap_genotypeTOP_20160222_full/out/PCA_data")
```

### Load distance matrix

```{r}
dist_matrix <- read.table("Data/ADAPTmap_genotypeTOP_20160222_full/out/PCA_data.mdist", header = F)
```
```{r}
# Extract breeds and individual names
fam_ind <- data.frame(
  famids = read.table("Data/ADAPTmap_genotypeTOP_20160222_full/out/PCA_data.mdist.id")[, 1],
  IID = read.table("Data/ADAPTmap_genotypeTOP_20160222_full/out/PCA_data.mdist.id")[, 2]
)
```

And Multidimensional scaling (MDS/PCoA) on the distance matrix of the dataset.
#### With `cmdscale()`
```{r}
mds_pop <- cmdscale(dist_matrix, eig = TRUE, 5)
head(mds_pop)
```

```{r}
nrow(fam)
nrow(famInd)
nrow(mds_pop$points)

eigenvec_populations <- cbind(fam_ind, mds_pop$points)
head(eigenvec_populations)
eigen_percent <- round(((mds_pop$eig) / sum(mds_pop$eig)) * 100, 1)

eigen_percent_goat_pop <-
  eigen_percent %>%
  as.data.frame(.) %>%
  rename("exp_var" = ".") %>%
  mutate("row" = as.numeric(rownames(.))) %>%
  select("row", "exp_var") %>%
  filter(row_number() <= 10) %>%
  as.tibble() %>%
  ggplot(aes(., x = row, y = exp_var)) +
  geom_col(color = "gray", fill = "gray", alpha = .75) +
  geom_line(color = "gray20") +
  geom_point(color = "grey20") +
  scale_x_continuous(expand = c(0.01, 0), breaks = c(1:10)) +
  scale_y_continuous(expand = c(0.01, 0)) +
  labs(title = "Explained variances in the first 10 dimensions", x = "Dimensions", y = "Explained variance (%)") +
  theme_classic() +
  theme(
    aspect.ratio = 1,
    panel.grid = element_blank(),
    axis.ticks = element_line()
  )
```
We have calculated the Eigenvectors and Eigenvalues.

```{r}
#library(ggpomological)

mds_goat <-
  eigenvec_populations %>%
  ggplot(.) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey20") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey20") +
  geom_point(aes(x = -`1`, y = `2`, color = famids), size = 1, alpha = .8, show.legend = FALSE) +
  labs(
    title = "MDS of worldwide goat population",
    x = paste0("Dimension 1 (", eigen_percent[1], " %)"),
    y = paste0("Dimension 2 (", eigen_percent[2], " %)")
  ) +
  #theme_pomological_fancy() +
  #scale_color_pomological() +
  theme_minimal() +
    theme(
      aspect.ratio = 1,
      panel.grid = element_blank(),
      axis.ticks = element_line()
    )

require("patchwork")
mds_goat_var <- mds_goat | eigen_percent_goat_pop


ggsave("Figures/goat_mds.pdf", plot = mds_goat_var, height = 5, width = 9, dpi = 600, device = grDevices::cairo_pdf)
ggsave("Figures/goat_mds.png", plot = mds_goat_var, height = 5, width = 9, dpi = 96)

```

## MDS in plink
There is an option to direct create the distance matrix, eigenvectors and eigenvalues directly in plink
```{r}
goats <- readr::read_tsv("Data/ADAPTmap_genotypeTOP_20160222_full/ADAPTmap_genotypeTOP_20160222_full.fam", col_names = F)
summary(goats)

selectedAnimals <- goats %>%
  filter(X1 == "ALP" | X1 == "ANG" | X1 == "BOE" | X1 == "CRE" | X1 == "BRK" | X1 == "RAN") %>%
  write_delim("Data/ADAPTmap_genotypeTOP_20160222_full/filtered_goats.txt", col_names = F)
```

```{r}
system(str_c(
  "plink \\
--bfile Data/ADAPTmap_genotypeTOP_20160222_full/ADAPTmap_genotypeTOP_20160222_full \\
--chr-set 29 no-xy \\
--autosome ",
  "--keep Data/ADAPTmap_genotypeTOP_20160222_full/filtered_goats.txt \\
             --nonfounders ",
  "--geno 0.1 \\
             --mind 0.1 \\
             --maf 0.05 ",
  "--make-bed \\
             --out Data/ADAPTmap_genotypeTOP_20160222_full/out/AfterQC_filtered_goats"
))
```

```{r}
system("plink \\
--bfile Data/ADAPTmap_genotypeTOP_20160222_full/out/AfterQC_filtered_goats \\
--chr-set 29 no-xy \\
--pca \\
--out Data/ADAPTmap_genotypeTOP_20160222_full/out/plinkMDS_filtered_goats")
```

Visualisation:
```{r}
eigenValues <- read_delim("Data/ADAPTmap_genotypeTOP_20160222_full/out/plinkMDS_filtered_goats.eigenval", delim = " ", col_names = F)
eigenVectors <- read_delim("Data/ADAPTmap_genotypeTOP_20160222_full/out/plinkMDS_filtered_goats.eigenvec", delim = " ", col_names = F)

## Proportion of variation captured by each vector
eigen_percent <- round((eigenValues / (sum(eigenValues))*100), 1)

# PCA plot
mds_gooat_filtered <-
  eigenVectors %>%
  mutate(
    X1A = case_when(
      X1 == "ALP" ~ "Alpine",
      X1 == "ANG" ~ "Angora",
      X1 == "BOE" ~ "Boer",
      X1 == "CRE" ~ "Creole",
      X1 == "BRK" ~ "Barki",
      X1 == "RAN" ~ "Rangeland"
  ),
    X1B = case_when(
      X1 == "ALP" ~ "Europe",
      X1 == "ANG" ~ "Worldwide",
      X1 == "BOE" ~ "Worldwide",
      X1 == "CRE" ~ "South America",
      X1 == "BRK" ~ "Africa",
      X1 == "RAN" ~ "Oceania"
    ))  %>%
ggplot(.) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey20") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey20") +
  geom_point(mapping = aes(x = X3, y = X4, color = X1A, shape = X1B), size = 2, alpha = .8, show.legend = TRUE ) +
    labs(
    title = "MDS of selected goat breeds",
    x = paste0("Dimension 1 (", eigen_percent[1,1], " %)"),
    y = paste0("Dimension 2 (", eigen_percent[2,1], " %)"),
    colour = "Goat breeds", shape = "Origin") +
  theme_minimal() +
    scale_color_frontiers() +
  theme(
      aspect.ratio = 1,
      panel.grid = element_blank(),
      axis.ticks = element_line()
    )

    mds_gooat_filtered

ggsave("Figures/goat_mds_filtered.pdf", plot = mds_gooat_filtered, height = 5, width = 6, dpi = 600, device = grDevices::cairo_pdf)
ggsave("Figures/goat_mds_filtered.png", plot = mds_gooat_filtered, height = 5, width = 6, dpi = 96)
knitr::plot_crop("Figures/goat_mds_filtered.pdf")
knitr::plot_crop("Figures/goat_mds_filtered.png")
```

```{r}
final <- (mds_goat) + (mds_gooat_filtered)

path <- "Figures/goat_mds_2.png"

ggsave(path, plot = final, width = 10, dpi = 120, type = "cairo")

knitr::plot_crop(path)
```
