---
title: "Wheat"
#author:
output:
  #rmdformats::html_clean:
  rmdformats::downcute:
    downcute_theme: "chaos"
    use_bookdown: false
    self_contained: true
    thumbnails: true
    lightbox: true
    highlight: tango
    gallery: true
    toc_depth: 3
    code_folding: show
    css: css/custom.css
knit:
  (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding =
  encoding, output_dir = "html") })
---

# Genomic project to investigate  the genetic basis of complex traits in Wheat (*T. aestivum*).

```{r}
rm(list = ls())
```

## Introduction

## Durum wheat (*Triticum turgidum* subsp. *durum*)

Data:
Dell'Acqua, Matteo (2022). Characterization data for the EtNAM population and Ethiopian durum wheat landraces diversity panel [Dataset]. Dryad. https://doi.org/10.5061/dryad.w6m905qrv form the orginal research article https://onlinelibrary.wiley.com/doi/10.1111/pbi.12538


Here we deal with data stored in the .Rdata format, great!
```{r}
data_et <- load("Data/Triticum-durum/etnam.data.gp.Rdata")
data_et

data <- load("Data/Triticum-durum/diversity.panel.data.gp.Rdata")
data

# Accessing the "met" object
met_data <- met

# Accessing the "farm" object
farm_data <- farm

# Accessing the "geno" object
geno_data <- geno

# Accessing the "snp.pos" object
snp_pos_data <- snp.pos

# Accessing the "info" object
info_data <- info
```